---
title: "The panelSEM Package"
author: "Christian Gische"
output: github_document
bibliography: man/bibliography.bib
csl: man/apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

## Installing the Package

The current version of the `panelSEM` package can be found on
[GitHub](https://github.com/christian-gische/panelSEM) and can be
installed directly via `RStudio` using the following code:[^1]

[^1]: <https://github.com/christian-gische/panelSEM>

```{r,  eval = FALSE}
# install the causalSEM package from GitHub
devtools::install_github("christian-gische/panelSEM")

# load lavaan and OpenMx
library(lavaan)
library(OpenMx)
```

## Purpose of the Package

The purpose of the package is to fit dynamic panel data models to data
and obtain parameter estimates. The class of models that can be
estimated ranges from simple linear models (e.g., the cross-lagged panel
model) to complex nonlinear nonadditive models with observed and
unobserved effect heterogeneity. The following figures display the
causal graphs of three prototypical models with increasing complexity.

### Linear Model without Unobserved Heterogeneity

The following figure displays the causal structure of a linear
cross-lagged panel model and is taken from @gische_beyond_2022.

```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}
\usetikzlibrary{arrows.meta}
\begin{tikzpicture}
\node (t1) at (8,0) {blood insulin};
\node (t2) at (8,-3) {blood glucose};   
             
\node (x1) at (0,0) {$X_{1}$};
\node (x3) at (3,0) {$X_{2}$};          
\node (x5) at (6,0) {$X_{3}$};

\node (x2) at (0,-3) {$Y_{1}$};
\node (x4) at (3,-3) {$Y_{2}$};
\node (x6) at (6,-3) {$Y_{3}$};

\draw[->] (x1) to node[above,near start,sloped]{\small $c_{xx}$} (x3); 
\draw[->] (x1) to node[above,near start,sloped]{\small $c_{yx}$} (x4);  
\draw[->] (x3) to node[above,near start,sloped]{\small $c_{xx}$} (x5);
\draw[->] (x3) to node[above,near start,sloped]{\small $c_{yx}$} (x6);\draw[->] (x2) to node[below,near start,sloped]{\small $c_{xy}$} (x3); 
\draw[->] (x2) to node[below,near start,sloped]{\small $c_{yy}$} (x4);  
\draw[->] (x4) to node[below,near start,sloped]{\small $c_{xy}$} (x5);
\draw[->] (x4) to node[below,near start,sloped]{\small $c_{yy}$} (x6);

\draw[<->,bend right=20,dashed] (x1) to node[left]{\small $\psi_{x_1y_1}$} (x2);
\draw[<->,bend left=60,dashed] (x1) to node[above]{\small $\psi_{x_1x_2}$} (x3);
\draw[<->,bend left=60,dashed] (x3) to node[above]{\small $\psi_{x_2x_3}$} (x5);
\draw[<->,bend right=60,dashed] (x2) to node[below]{\small $\psi_{y_1y_2}$} (x4);
\draw[<->,bend right=60,dashed] (x4) to node[below]{\small $\psi_{y_2y_3}$} (x6);

\draw[<->,bend right=20,dashed] (x3) to node[right]{\small $\psi_{xy}$} (x4);
\draw[<->,bend right=20,dashed] (x5) to node[right]{\small $\psi_{xy}$} (x6);

\end{tikzpicture}
```

### Linear Model with Additive Unobserved Heterogeneity

The following figure displays the causal structure of a dynamic panel
data model with additive heterogeneity and is taken from
@gische_west_voelkle_2021.

```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}
\usetikzlibrary{arrows.meta}
\begin{tikzpicture}[latent/.style={ellipse,draw=black!80,dashed,thick,inner sep=2,text centered}]
     
\node (x1) at (0,0) {$X_{1}$};
\node (x2) at (3,0) {$X_{2}$};          
\node (x3) at (6,0) {$X_{3}$};
\node (x4) at (9,0) {$X_{4}$};
    
\node (y1) at (0,-3) {$Y_{1}$};
\node (y2) at (3,-3) {$Y_{2}$};
\node (y3) at (6,-3) {$Y_{3}$};
\node (y4) at (9,-3) {$Y_{4}$};

\draw[->] (x1) to node[above,midway,sloped]{} (x2); 
\draw[->] (x1) to node[above,near start,sloped]{} (y2); 
\draw[->] (x2) to node[above,midway,sloped]{} (x3);
\draw[->] (x2) to node[above,near start,sloped]{} (y3); 
\draw[->] (x3) to node[above,near start,sloped]{} (x4);
\draw[->] (x3) to node[above,near start,sloped]{} (y4);
\draw[->] (y1) to node[above, near start,sloped]{}(x2); 
\draw[->] (y1) to node[below,midway,sloped]{}(y2);  
\draw[->] (y2) to node[above, near start,sloped]{} (x3);
\draw[->] (y2) to node[below,midway,sloped]{} (y3);
\draw[->] (y3) to node[above, near start,sloped]{} (x4);
\draw[->] (y3) to node[below,near start,sloped]{} (y4);

\draw[<->,bend right,dashed] (x1) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (y1);

\node[latent] (etax1) at (4,2) {$\eta_{x}$};
\draw[->] (etax1) to node[above, midway,sloped]{} (x1);     
\draw[->] (etax1) to node[left, near start]{}(x2); 
\draw[->] (etax1) to node[left, near start]{}(x3); 
\draw[->] (etax1) to node[below, near start]{}(x4);
\draw [->, bend right=90,out=-50, looseness=1.5] (etax1) to node[above, near start,sloped]{} (y1);

\node[latent] (etay0) at (4,-5) {$\eta_{y}$};
\draw[->] (etay0) to node[below, near start,sloped]{} (y1);     
\draw[->] (etay0) to node[left, near start]{} (y2); 
\draw[->] (etay0) to node[above, near start]{} (y3);
\draw[->] (etay0) to node[below, near start]{} (y4);
\draw [->, bend left=90, out=50, looseness=1.5] (etay0) to node[below, near start,sloped]{} (x1);

\draw[<->,bend left,dashed, out=90, in=90, looseness=3] (etax1) to (etay0);

\end{tikzpicture}
```

### Nonlinear Model with Nonadditive Unobserved Heterogeneity

The following figure displays the causal structure of a nonlinear
dynamic panel data model with nonadditive heterogeneity and is taken
from @gische_west_voelkle_2023

```{tikz, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}
\usetikzlibrary{arrows.meta}
\begin{tikzpicture}[latent/.style={ellipse,draw=black!80,dashed,thick,inner sep=2,text centered}]
             
\node (x1) at (0,0) {$X_{1}$};
\node (x2) at (3,0) {$X_{2}$};          
\node (x3) at (6,0) {$X_{3}$};
\node (x4) at (9,0) {$X_{4}$};

\node (y1) at (0,-3) {$Y_{1}$};
\node (y2) at (3,-3) {$Y_{2}$};
\node (y3) at (6,-3) {$Y_{3}$};
\node (y4) at (9,-3) {$Y_{4}$};

\node (zx) at (-2,2) {$Z_1$};
\draw[->,gray!70] (zx) to node[above,midway]{} (x1); 
\draw[->,gray!70] (zx) to node[above,midway]{} (x2); 
\draw[->,gray!70] (zx) to node[above,midway]{} (x3);
\draw[->,gray!70] (zx) to node[above,midway]{} (x4);


\node (zxy) at (-2,-1.5) {$Z_2$};
\draw[->,gray!70] (zxy) to node[above,midway]{} (x1); 
\draw[->,gray!70] (zxy) to node[above,midway]{} (x2); 
\draw[->,gray!70] (zxy) to node[above,midway]{} (x3);
\draw[->,gray!70] (zxy) to node[above,midway]{} (x4);
\draw[->,gray!70] (zxy) to node[above,midway]{} (y1); 
\draw[->,gray!70] (zxy) to node[above,midway]{} (y2); 
\draw[->,gray!70] (zxy) to node[above,midway]{} (y3);
\draw[->,gray!70] (zxy) to node[above,midway]{} (y4);



\node (zy) at (-2,-5) {$Z_3$};
    
\draw[->,gray!70] (zy) to node[above,midway]{} (y1); 
\draw[->,gray!70] (zy) to node[above,midway]{} (y2); 
\draw[->,gray!70] (zy) to node[above,midway]{} (y3);
\draw[->,gray!70] (zy) to node[above,midway]{} (y4);

\draw[<->,bend right,dashed,gray!70] (zx) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (zxy);
\draw[<->,bend right,dashed,gray!70] (zx) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (zy);
\draw[<->,bend right,dashed,gray!70] (zxy) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (zy);

\draw[->,gray!70] (zx) to node[above,midway]{} (y1);
\draw[->,gray!70] (zy) to node[above,midway]{} (x1);
 
\draw[->] (x1) to node[above,midway,sloped]{} (x2); %\scriptsize $c_{x_2x_1}$
\draw[->] (x1) to node[above,near start,sloped]{} (y2); 
\draw[->] (x2) to node[above,midway,sloped]{} (x3); % \scriptsize $c_{x_3x_2}$
\draw[->] (x2) to node[above,near start,sloped]{} (y3); 
\draw[->] (x3) to node[above,near start,sloped]{} (x4); %\scriptsize $c_{x_4x_3}$
\draw[->] (x3) to node[above,near start,sloped]{} (y4);
\draw[->] (y1) to node[above, very near start,sloped]{}(x2); %\scriptsize $c_{x_2y_1}$
\draw[->] (y1) to node[below,midway,sloped]{}(y2);  % \scriptsize $c_{y_2y_1}$
\draw[->] (y2) to node[above, near start,sloped]{} (x3); %\scriptsize $c_{x_3y_2}$
\draw[->] (y2) to node[below,midway,sloped]{} (y3); %\scriptsize $c_{y_3y_2}$
\draw[->] (y3) to node[above, near start,sloped]{} (x4); %\scriptsize $c_{x_4y_3}$
\draw[->] (y3) to node[below,near start,sloped]{} (y4); %\scriptsize $c_{y_4y_3}$

\draw[<->,bend right,dashed] (x1) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (y1);


\node[latent] (etax0) at (2,2) {$\eta_{x}$};

\draw[->] (etax0) to node[above, midway,sloped]{} (x1); 
\draw[->] (etax0) to node[left, near start,gray!70]{}(x2); 
\draw[->] (etax0) to node[left, near start,gray!70]{}(x3); 
\draw[->] (etax0) to node[below, near start,gray!70]{}(x4);


\node[latent] (etay0) at (2,-5) {$\eta_{y}$};

\draw[->] (etay0) to node[below, near start,sloped]{} (y1);     
\draw[->] (etay0) to node[left, near start, gray!70]{}  (y2); 
\draw[->] (etay0) to node[above, near start, gray!70]{} (y3);
\draw[->] (etay0) to node[below, near start, gray!70]{} (y4);

\node[latent] (etax1) at (6,2) {$\eta_{xy}$};

\draw[->] (etax1) to node[above,midway]{} (x1);
\draw[->] (etax1) to node[above,midway]{} (x2); 
\draw[->] (etax1) to node[above,midway]{} (x3);
\draw[->] (etax1) to node[above,midway]{} (x4);

\node[latent] (etay1) at (6,-5) {$\eta_{yx}$};

\draw[->] (etay1) to node[above,midway]{} (y1);
\draw[->] (etay1) to node[above,midway]{} (y2); 
\draw[->] (etay1) to node[above,midway]{} (y3);
\draw[->] (etay1) to node[above,midway]{} (y4);


\draw[<->,bend right,dashed,looseness=0.5] (etay0) to node[left,sloped,rotate=180]{} (etay1);
\draw[<->,bend left,dashed,looseness=0.5] (etax0) to node[left,sloped,rotate=180]{} (etax1);
\draw[<->,bend left=120, dashed, looseness=4.5] (etax0) to (etay0);
\draw[<->,bend left=80,dashed, looseness=1.8] (etax1) to (etay1);
\draw[<->,bend left=80, out=100, dashed, looseness=3] (etax0) to (etay1);
\draw[<->,bend right=80, out=-100, dashed, looseness=3] (etay0) to (etax1);




\draw[<->,bend right,dashed] (x1) to node[left,sloped,rotate=180,yshift=10pt,xshift=25pt]{} (y1);
\draw [->, bend right=90,out=-80, looseness=1.5] (etax0) to node[above, near start,sloped]{} (y1);
\draw [->, bend right=90,out=-80, looseness=1.6] (etax1) to node[above, near start,sloped]{} (y1);
\draw [->, bend left=90, out=80, looseness=1.5] (etay0) to node[below, near start,sloped]{} (x1);
\draw [->, bend left=90, out=90, looseness=1.6] (etay1) to node[below, near start,sloped]{} (x1);   
\end{tikzpicture}
```

## The `fit_panel_sem`-Function

The central function of the package is the `fit_panel_sem`-function
which allows users to (i) specify a model, (ii) provide the data, and
(iii) fit the model to the data. The following code chunk displays a
minimal example of a function when only mandatory arguments are
specified.

```{r message=FALSE, eval=FALSE, warning=FALSE}
panelSEM_object <- fit_panel_sem(data = data,
                                 time_varying_variables = time_varying_variables,
                                 time_invariant_variables = time_invariant_variables,
                                 linear = linear,
                                 heterogeneity = heterogeneity)
```

The mandatory arguments are listed below and will be explained in detail
throughout the subsequent sections:

- `data`: the data set    
- `time_varying_variables`: a list of time-varying observed variables 
- `time_invariant_variables`: a list of time-invariant observed covariates      
- `linear`: an argument specifying if the model is linear or nonlinear    
- `heterogeneity`: an argument specifying the type of unobserved heterogeneity


### The `data` argument

The data set needs to be provided as a `.txt`-file and organized in the
wide format. 

### The `time_varying_variables` argument  

### The `time_invariant_variables` argument  

### The `linear` and the `unobserved_heterogeneity` argument  

```{tikz linear_heterogeneity, fig.ext = 'png', echo = FALSE}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{fit}
\usetikzlibrary{backgrounds}
\usetikzlibrary{arrows.meta}
\begin{tikzpicture}[align=center,
box/.style={draw,rectangle,minimum width = 3.5cm, minimum height= 1.5cm, text width=3.5cm, align=left}]
\matrix (conmat) [row sep=.1cm,column sep=.1cm] {
    \node (tpos) [box,
    label=left:\texttt{linear == TRUE},
    label=above:\texttt{"homogeneous"},
    ] {observed variables enter equation linear \\\vspace{16pt} no unobserved heterogeneity (no $\eta$-terms)\\\vspace{16pt} covariances across occasions and within occasions (see Linear Model without Unobserved Heterogeneity)};
    &
    \node (fneg) [box,
    label=above:\texttt{"additive"},
    ] {observed variables enter equation linear \\\vspace{16pt} additive unobserved heterogeneity (additive $\eta$-terms)};
    &
    \node (fneg1) [box,
    label=above:\texttt{"cross-lagged"}
    ] {observed variables enter equation linear \\\vspace{3pt} nonadditive unobserved heterogeneity (multiplicative $\eta$-terms)};   
    \\
    \node (fpos) [box,
    label=left:\texttt{linear == FALSE}] 
    {observed variables enter equation nonlinear (product terms) \\\vspace{16pt} no unobserved heterogeneity (no $\eta$-terms)};
    &
    \node (tneg) [box,
    ] {observed variables enter equation nonlinear (product terms)\\\vspace{16pt} additive unobserved heterogeneity (additive $\eta$-terms)};
    &
    \node (tneg1) 
    [box,
    ] {observed variables enter equation nonlinear (product terms)\\\vspace{3pt} nonadditive unobserved heterogeneity (multiplicative $\eta$-terms)};
    \\
};
%\node [left=.05cm of conmat,anchor=center,text width=1.5cm] {\textbf{average vs. \\ person-specific}};
%\node [above=.05cm of conmat] {\textbf{forecast vs. prediction}};
\end{tikzpicture}
```

Additionally to the basic combinations displayed in the figure above, a model can exhibit more than one type of unobserved heterogeneity. Thus, the argument `heterogeneity` can take the value  `heterogeneity == c("additive","cross-lagged")`. We now illustrate the different types of equations.

### Linear, homogeneous

$$
Y_{t} = c_{yx}X_{t-1}+c_{yy}Y_{t-1}+c_{yz_{2}}Z_2+c_{yz_{3}}Z_3+\varepsilon_{yt}
$$

The equation is linear in the observed variables (because all observed variables are enter the equation in a linear way) and homogeneous (there is no unobserved person-specific heterogeneity; there are no $\eta$-terms).

### Linear, additive heterogeneity

$$
Y_{t} = c_{yx}X_{t-1}+c_{yy}Y_{t-1}+c_{yz_{2}}Z_2+c_{yz_{3}}Z_3+\eta_y+\varepsilon_{yt}
$$

The equation is linear in the observed variables (because all observed variables are enter the equation in a linear way) and the unobserved heterogeneity is additive (because there the $\eta$-term enters the equation in an additive way).

### Nonlinear, homogeneous

$$
Y_{t} = (c_{yx} + c_{yxz_2}Z_2 + c_{yxz_3}Z_3)X_{t-1}+c_{yy}Y_{t-1}+c_{yz_2}Z_2+c_{yz_3}Z_3+\varepsilon_{yt}
$$

The equation is nonlinear in the observed variables (because there are multiplicative terms of observed variables, namely $Z_2X_{t-1}$) and homogeneous (there is no unobserved person-specific heterogeneity; there are no $\eta$-terms).

### Nonlinear, additive heterogeneity

$$
Y_{t} = (c_{yx} + c_{yxz_2}Z_2 + c_{yxz_3}Z_3)X_{t-1}+c_{yy}Y_{t-1}+c_{yz_2}Z_2+c_{yz_3}Z_3+\eta_y+\varepsilon_{yt}
$$

The equation is nonlinear in the observed variables (because there are multiplicative terms of observed variables, namely $Z_2X_{t-1}$) and the unobserved heterogeneity is additive (because there the $\eta$-term enters the equation in an additive way).

### Nonlinear, additive  and cross-lagged heterogeneity

$$
Y_{t} = (c_{yx} + c_{yxz_2}Z_2 + c_{yxz_3}Z_3)X_{t-1}+c_{yy}Y_{t-1}+c_{yz_2}Z_2+c_{yz_3}Z_3+\eta_{y}+\eta_{yx}X_{t-1}+\varepsilon_{yt}
$$

The equation is nonlinear in the observed variables (because there are multiplicative terms of observed variables, namely $Z_2X_{t-1}$) and the unobserved heterogeneity is both additive (because the term $\eta_y$ enters the equation in an additive way) and there is also unobserved heterogeneity in the cross-lagged coefficient (because the term $\eta_{yx}$ enters the equation in a multiplicative way, namely $\eta_{yx}X_{t-1}$).


## References  

